Constraint Grammar as a SAT problem

### 2015-03-30 paper/nodalida2015.tex
more text

\author{Inari Listenmaa \and Koen Claessen \\
  Chalmers University of Technology, Gothenburg, Sweden \\
  {\tt \{inari,koen\}@chalmers.se} }

@@

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
Applying logic to reductionist grammars has been explored earlier by 
\cite{lager98,lager_nivre01}, but it was never adopted for use---logic programming was simply 
too slow to be used for tagging or parsing. 
However, modern SAT solvers are fast and used for many applications where constraints must be
satisfied, such as software and hardware verification.
%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.

It is attractive to model CG in terms of logic:
these techniques are well-studied, and they capture more dependencies between the tags than the standard CG.

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of ``last'' is not applicable.
With these constraints, we get two solutions: \texttt{det, n} and \texttt{prn, v pri p3, v imp p2}. 
The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} 
for free, and ordering of the rules doesn't matter. 
%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

The worst-case complexity of SAT is exponential, whereas the standard implementations 
of CG are polynomial, but with advances in SAT solving techniques, the performance 
in the average case is much more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
%We do foreach sentence: foreach rule; 
%whereas VISL-CG3 has foreach rule: foreach sentence.
%the size of the SAT problem is dependent of the number of rules

Table~\ref{table:time} shows the execution time compared to VISL-CG3. 
The number of rules in the grammar is 
Maximisation is the most costly operation
Also, some of the time is explained simply by
SATCG is just a first proof-of-concept implementation and can be optimised.

\begin{table}
  \centering
  \begin{tabular}{|l|l|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG} & \textbf{VISL-CG3} \\ \hline
              153,386    & 99    & 55.7s     & 4.8s \\  \hline	                         
                186,460  & 19    &  13.7s      & 4.2s\\ %\hline
                186,460  & 130  &  1m14.1s & 6.1s \\ %\hline
                186,460  & 261  &  2m53.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\paragraph{Performance against VISL-CG3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all ambiguities, and ran both VISL-CG3 and our implementation with the same grammar. 
Treating the chosen tagset as the gold standard, the disambiguation by SATCG achieves 75.09 \% correct result and VISL-CG3 75.80 \%.

% \paragraph{Removing rules}
% \begin{table}
%   \centering
%   \begin{tabular}{|l|c|c|}
%      \hline
%     \textbf{\# rules} & \textbf{SATCG} & \textbf{VISL-CG3} \\
%     \hline 
% 	19 &  83.15 & 83.33 \\ \hline  % 19 out of 23 preselected by authors
% 	18 &  83.19 & 83.04 \\ \hline  % 18 out of 19
% 	17 &  83.201 & 83.03 \\ \hline   % 17 out of 18/19, no diff
% 	16 &  83.197 & 83.030 \\ \hline  % 16 out of 17/18, no diff
% 	15 &  83.188 & 83.022\\ \hline  % 15 out of 16/17. Tie with 14 for SATCG, chose to retain the rule that gave better results for VISL-CG3.
% 	14 &  83.180 & 83.005 \\ \hline  % Tie with 15 for SATCG.
% 	13 &  83.159 & 82.951 \\ \hline  % 13 out of 15 
% 	12 &  83.134 & 82.884\\ \hline  % 12 out of 14
% 	11 &  83.089 & 82.847\\ \hline  
% 	10 &  83.051 & 82.802\\ \hline  
% 	9  & 82.997 & 82.802 \\ \hline
% 	8  & 82.906 & 82.681 \\ \hline  % REMOVE:r_imp Imp IF (0 Subj) (-1* CnjSub) ; 
% 	7  & 82.743 & 82.594 \\ \hline  % SELECT:s_pro PrnIndep IF (1C VerbFin) ;
% 	6  & 82.556 & 82.398\\ \hline  %  REMOVE:r_v_v Verb IF (-1C Vblex) ;
% 	5  & 82.327 & 82.028 \\ \hline % SELECT:s_adv Adv IF (1 Cnj_Rel_End OR Prep OR Det)
% 	4  & 82.065 & 82.028 \\ \hline % REMOVE:r_vfin_prep VerbFin IF (0 Prep) (1 Inf) 
% 	3  & 81.387 & 81.233 \\ \hline % REMOVE:r_noun N    IF (-1 N OR Prn)
% 	2  & 80.23 & 79.97 \\ \hline % SELECT:s_pri_1 (pri p3) IF (0 (vblex imp p2 sg))
% 	1  & 76.69 & 76.62 \\ \hline 
% 	0  & 70.70 & 70.70 \\ \hline
%   \end{tabular}
%   \caption{Result of removing rules one by one, and how close to the gold standard they get.}
%   \label{table:results_remove}
% \end{table}
% In this experiment, we took 19 rules written by the authors, which give 83.15 \% correct results with SATCG and 83.33 \% with VISL-CG3.
% Then we removed rules one by one; after getting a score for $n$ rules, we took the combination of $n-1$ rules for which SATCG gives the best score, and we ran the rule set with VISL-CG3 for comparison. Table~\ref{table:results_remove} shows the percentages compared to the gold standard for both systems.

>>>>

% \author{Inari Listenmaa \and Koen Claessen \\
%  Chalmers University of Technology, Gothenburg, Sweden \\
%  {\tt \{inari,koen\}@chalmers.se} }

@@

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.

Formal logic is well-studied and has an old tradition.
The rules encoded in logic capture richer dependencies between the
tags than normal reductionist grammars.

Applying logic to reductionist grammars has been explored earlier by 
\cite{lager98,lager_nivre01}, but it was never adopted for use.
% ; logic programming was too slow to be used for tagging or parsing. 
Since those works, SAT solving techniques have improved significantly, and 
they are used in domains such as microprocessor design and computational 
biology---these problems easily match or exceed CG in complexity. 
\todo{is this true for grammars with thousands of rules?}

Due to these advances, we were able to revisit the idea and develop it further. 
% Whereas \cite{lager98} provides a proof-of-concept description, 
Missing from the previous logic-based works, 
we solve eventual rule conflicts by finding a solution that 
discards the least number of rules.
% allows the maximal number of rules to be applied.
We test our implementation by parsing texts in the order of 10,000s--100,000s
words, using grammars with hundreds of rules.

%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of ``last'' is not applicable.
With these constraints, we get two solutions: 
\begin{enumerate}
\item []
\begin{verbatim}
"<la>"
        "el" det def f sg
"<casa>"
        "casa" n f sg
\end{verbatim}
\item [and]
\begin{verbatim}
"<la>"
        "lo" prn p3 f sg
"<casa>"
        "casar" v pri p3 sg
        "casar" v imp p2 sg
\end{verbatim}
\end{enumerate} 
% \texttt{det, n} and \texttt{prn, v pri p3, v imp p2}. 
The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} 
for free, and ordering of the rules doesn't matter. 
%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

The worst-case complexity of SAT is exponential, whereas the standard implementations 
of CG are polynomial, but with advances in SAT solving techniques, the performance 
in the average case is much more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
%We do foreach sentence: foreach rule; 
%whereas VISL CG-3 has foreach rule: foreach sentence.
%the size of the SAT problem is dependent of the number of rules

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is 
Maximisation is the most costly operation
Also, some of the time is explained simply by
SATCG is just a first proof-of-concept implementation and can be optimised.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG} & \textbf{VISL CG-3} \\ \hline
               % 124,493  & 99    & 55.7s   & 4.8s \\  \hline	                         
                384,155  & 19    &  13.7s  & 4.2s\\ %\hline
                384,155  & 130  &  1m14.1s & 6.1s \\ %\hline
                384,155  & 261  &  2m53.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\paragraph{Performance against VISL CG-3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all ambiguities, and ran both VISL CG-3 and our implementation with the same grammar. 
Treating the chosen tagset as the gold standard, the disambiguation by SATCG achieves 75.09 \% correct result and VISL CG-3 75.80 \%.

% \paragraph{Removing rules}
% \begin{table}
%   \centering
%   \begin{tabular}{|l|c|c|}
%      \hline
%     \textbf{\# rules} & \textbf{SATCG} & \textbf{VISL CG-3} \\
%     \hline 
% 	19 &  83.15 & 83.33 \\ \hline  % 19 out of 23 preselected by authors
% 	18 &  83.19 & 83.04 \\ \hline  % 18 out of 19
% 	17 &  83.201 & 83.03 \\ \hline   % 17 out of 18/19, no diff
% 	16 &  83.197 & 83.030 \\ \hline  % 16 out of 17/18, no diff
% 	15 &  83.188 & 83.022\\ \hline  % 15 out of 16/17. Tie with 14 for SATCG, chose to retain the rule that gave better results for VISL CG-3.
% 	14 &  83.180 & 83.005 \\ \hline  % Tie with 15 for SATCG.
% 	13 &  83.159 & 82.951 \\ \hline  % 13 out of 15 
% 	12 &  83.134 & 82.884\\ \hline  % 12 out of 14
% 	11 &  83.089 & 82.847\\ \hline  
% 	10 &  83.051 & 82.802\\ \hline  
% 	9  & 82.997 & 82.802 \\ \hline
% 	8  & 82.906 & 82.681 \\ \hline  % REMOVE:r_imp Imp IF (0 Subj) (-1* CnjSub) ; 
% 	7  & 82.743 & 82.594 \\ \hline  % SELECT:s_pro PrnIndep IF (1C VerbFin) ;
% 	6  & 82.556 & 82.398\\ \hline  %  REMOVE:r_v_v Verb IF (-1C Vblex) ;
% 	5  & 82.327 & 82.028 \\ \hline % SELECT:s_adv Adv IF (1 Cnj_Rel_End OR Prep OR Det)
% 	4  & 82.065 & 82.028 \\ \hline % REMOVE:r_vfin_prep VerbFin IF (0 Prep) (1 Inf) 
% 	3  & 81.387 & 81.233 \\ \hline % REMOVE:r_noun N    IF (-1 N OR Prn)
% 	2  & 80.23 & 79.97 \\ \hline % SELECT:s_pri_1 (pri p3) IF (0 (vblex imp p2 sg))
% 	1  & 76.69 & 76.62 \\ \hline 
% 	0  & 70.70 & 70.70 \\ \hline
%   \end{tabular}
%   \caption{Result of removing rules one by one, and how close to the gold standard they get.}
%   \label{table:results_remove}
% \end{table}
% In this experiment, we took 19 rules written by the authors, which give 83.15 \% correct results with SATCG and 83.33 \% with VISL CG-3.
% Then we removed rules one by one; after getting a score for $n$ rules, we took the combination of $n-1$ rules for which SATCG gives the best score, and we ran the rule set with VISL CG-3 for comparison. Table~\ref{table:results_remove} shows the percentages compared to the gold standard for both systems.
